---
title: "Assignment1 Data607"
author: "Madina Kudanova"
format: html
editor: visual
---

## Introduction

This data set contains information on popular baby names in NYC, including birth year, ethnicity, gender and popularity rank. The data set is useful for exploring trends in baby names among New Yorkers. The original data set sourced from Data.gov: <https://catalog.data.gov/dataset/popular-baby-names>

## Approach

First, the data set will be saved on my GitHub repository (DATA607) to ensure reproducibility. Then, the raw data will be loaded into R and reviewed. Here one variable ("Ethnicity") will be removed, and another will be renamed from "Rank" to "Popularity Rank" in order to improve clarity. These steps are intended to produce a more focused data frame.

## Code Base

1\) Loading and displaying data set

```{r}
URL <-"https://raw.githubusercontent.com/MKudanova/Data607/refs/heads/main/Assignment1/Popular_Baby_Names.csv"

babies_raw <- read.csv(URL)
head(babies_raw)
print(head(babies_raw))
```

2\) Transforming data set

```{r}
babies_new <- babies_raw[, c("Year.of.Birth", "Gender", "Child.s.First.Name", "Count", "Rank") ]
names(babies_new) <- c("Year_of_Birth", "Gender", "First_Name", "Count", "Popularity_Rank")
head(babies_new)

```

## Conclusion

As my first assignment using the R programming language and the Quarto format, this project 
provided an opportunity to practice loading a dataset and performing basic data transformations 
in a reproducible environment. Through this process, a subset of relevant variables was selected 
and a column name was renamed to improve clarity. This assignment established a foundation for 
working with data in R. As a next step, this cleaned data set could be extended by filtering 
to a specific year and summarizing name popularity using the Count and Popularity_Rank variables.
