---
title: "NY Times API"
author: "Madina KUdanova"
format: html
editor: visual
---

## Approach

I will use the New York Times Article Search API to analyze how frequently different newspaper sections publish articles on a specific topic. The ultimate goal is to answer the question: which New York Times sections published the most articles about artificial intelligence in 2025, and how this distribution compares to 10 years earlier.

To answer this question, I will first obtain access to the API by creating a New York Times developer account and securely storing my API key as an environment variable using Sys.getenv("NYT_API_KEY"). I will then construct requests to the Article Search API endpoint using a keyword query (“AI” and "artificial intelligence) and two defined date ranges: January 1 to December 31, 2025, and January 1 to December 31, 2015. I will include the section_name facet parameter to summarize article counts by section for each time period.

Once the requests are made, I will parse the returned JSON responses using R, specifically with the jsonlite package, and inspect their structure to identify the relevant fields. Because the API response is nested, I will extract the section-level counts from the facets portion of each response and transform them into tidy data frames. Each dataset will contain section names and their corresponding article counts.

The datasets for both years will then be combined into a single structure that includes a year variable, allowing for direct comparison. Finally, I will analyze the data by comparing article counts across sections between 2015 and 2025 to identify shifts in coverage over time. The final Quarto document will include a description of the selected API and endpoint, the parameters used in the requests, the R code for authentication and data retrieval, the resulting tidy data frames, and a brief explanation of any data-cleaning decisions made during the process.

## Code Base

```{r message=FALSE}
library(tidyverse)
library(httr2)

```

```{r message=FALSE}
library(jsonlite)
```

## API Key Authentication

The API key is stored securely as an environment variable and accessed with `Sys.getenv()`.

To set it once on your computer, you can run something like this in your `.Renviron` file:

`NYT_API_KEY=your_actual_key_here`

Then restart R.

```{r}
readRenviron("~/.Renviron")
nyt_key <- Sys.getenv("NYT_API_KEY")

if (nyt_key == "") {
  stop("NYT_API_KEY not found. Please store your API key in an environment variable.")
}
```

```{r}
req <- request("https://api.nytimes.com/svc/search/v2/articlesearch.json") |>
  req_url_query(
    q = "artificial intelligence OR AI OR machine learning OR language model",
    begin_date = "20250101",
    end_date = "20251231",
    page = 0,
    `api-key` = nyt_key
  )

resp <- req_perform(req)
resp

```

```{r}
# convert JSON response into R list
nyt_data <- resp_body_json(resp, simplifyVector = TRUE)

# inspect structure
names(nyt_data)
names(nyt_data$response)
```

```{r}
# create a tidy dataframe from the JSON response
articles_df <- as_tibble(nyt_data$response$docs) |>
  transmute(
    headline = headline$main,  # extract main headline text
    section_name = if_else(is.na(section_name) | section_name == "", "Unknown", section_name),  # fill missing sections
    pub_date = as.Date(substr(pub_date, 1, 10)),  # convert publication date
    web_url,          # article link
    news_desk,        # newsroom category
    document_type     # type of article
  )

# view the cleaned dataframe
articles_df
```

```{r}
# count articles by section
section_counts <- articles_df |>
  count(section_name, sort = TRUE)

section_counts
```

```{r}
# plot top sections
section_counts |>
  slice_max(order_by = n, n = 10) |>
  ggplot(aes(x = reorder(section_name, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top NYT Sections for AI Articles in 2025",
    x = "Section",
    y = "Article Count"
  )
```

```{r}
# request 2015 data (1 page to avoid rate limits)
req_2015 <- request("https://api.nytimes.com/svc/search/v2/articlesearch.json") |>
  req_url_query(
    q = "artificial intelligence OR AI OR machine learning",
    begin_date = "20150101",
    end_date = "20151231",
    page = 0,
    `api-key` = nyt_key
  )

resp_2015 <- req_perform(req_2015)

# parse JSON
nyt_data_2015 <- resp_body_json(resp_2015, simplifyVector = TRUE)

# create dataframe for 2015
articles_2015 <- as_tibble(nyt_data_2015$response$docs) |>
  transmute(
    headline = headline$main,
    section_name = if_else(is.na(section_name) | section_name == "", "Unknown", section_name),
    pub_date = as.Date(substr(pub_date, 1, 10)),
    web_url,
    news_desk,
    document_type
  )

articles_2015
```

```{r}
# add year labels
nyt_2025 <- articles_df |> mutate(year = 2025)
nyt_2015 <- articles_2015 |> mutate(year = 2015)

# combine both datasets
nyt_all <- bind_rows(nyt_2015, nyt_2025)

nyt_all
```

```{r}
# count by year and section
section_compare <- nyt_all |>
  count(year, section_name, sort = TRUE)

section_compare
```

```{r}
ggplot(section_compare, aes(x = n, y = reorder(section_name, n))) +
  geom_col() +
  facet_wrap(~ year, scales = "free_y") +
  labs(
    title = "NYT Sections for AI Articles: 2015 vs 2025",
    x = "Article Count",
    y = "Section"
  )
```

```{r}
section_totals <- section_compare |>
  group_by(year) |>
  summarize(total = sum(n))

section_totals
section_pct <- section_compare |>
  group_by(year) |>
  mutate(
    total = sum(n),
    pct = (n / total) * 100
  ) |>
  ungroup()

section_pct
```

## Conclusion

The comparison between 2015 and 2025 shows a clear shift in how artificial intelligence is covered across New York Times sections. In 2015, coverage was concentrated primarily in Technology and Science, suggesting a more technical and research-oriented focus.

By 2025, AI-related articles appear across a broader range of sections, including Technology, Business, Opinion, and others, indicating that AI has expanded beyond a specialized topic into wider industry and public discourse.

A percentage-based comparison highlights this shift more clearly. The Business section shows the largest increase, growing from 0% in 2015 to 20% in 2025, reflecting the growing economic relevance of AI. In contrast, Science declines from 20% to 0%, suggesting reduced emphasis on AI as a purely research-driven subject. Technology remains the dominant section in both years, increasing from 40% to 50%, while the appearance of new sections in 2025 reflects the broader integration of AI across different domains.

Because this analysis is based on the first page of API results for each year, the findings should be interpreted as a sample rather than a complete count of all published articles.
