---
title: "Laptop Price"
author: "Madina Kudanova"
format: html
editor: visual
---

## Introduction

This project develops practical competency in transforming wide-format datasets into tidy formats suitable for downstream analysis. All transformations and data manipulations are performed using the \`tidyr\` and \`dplyr\` packages in R, with visualizations built using \`ggplot2\`.

**Dataset used:**

**Uncleaned Laptop Price Dataset**: https://www.kaggle.com/datasets/ehtishamsadiq/uncleaned-laptop-price-dataset

## Approach

This dataset contains 1,303 laptop models covering specs like CPU, RAM, storage, GPU, screen resolution, OS, weight, and price. As the name suggests, the data comes as-is — messy, mixed, and in need of some serious cleaning before anything useful can be done with it.

The first thing I'll need to tackle is all the columns that are stuffed with multiple pieces of information. For example, RAM is stored as "8GB" instead of just 8, the CPU column bundles brand, model, and clock speed together, and Memory mixes storage size with storage type. I'll use \`stringr\` to pull out the numeric values and separate the meaningful parts into their own columns.

Once the data is clean, I'll reshape the specification columns into long format using \`pivot_longer()\`, which will make it much easier to compare specs across laptop types. From there I want to explore which laptop types give you the most RAM and storage for your money — essentially a value-for-price analysis, using laptop type as the main grouping variable rather than brand, since type is a better reflection of what you're actually getting.

## Base Code

```{r}
#| message: false
library(tidyverse)
```

## Data Import

```{r}
# Raw Data Import from my GitHub repository 
raw_data <- read_csv(
  "https://raw.githubusercontent.com/MKudanova/Data607/refs/heads/main/Project2/laptopData.csv")
glimpse(raw_data)
head(raw_data)
```

## Data Cleaning

```{r}
#| warning: false

library(tidyverse)

# Clean and standardize the raw laptop dataset
clean_data <- raw_data |>
  # Remove unnecessary index column if present
  select(-`Unnamed: 0`) |>
  
  # Standardize column names to lowercase
  rename_with(tolower) |>
  
  # Trim whitespace and convert empty strings to NA in all character columns
  mutate(
    across(where(is.character), ~ na_if(str_trim(.x), "")),
    
    # Convert obvious placeholder values to NA
    across(c(memory, ram, weight, inches, opsys), ~ na_if(.x, "?"))
  ) |>
  
  # Clean and transform variables
  mutate(
    # Convert embedded numeric text to numeric values
    ram = parse_number(ram),
    weight = parse_number(weight),
    inches = parse_number(inches),
    
    # Extract storage size in GB
    storage_gb = parse_number(memory),
    
    # Extract main storage type
    storage_type = case_when(
      str_detect(memory, regex("ssd", ignore_case = TRUE)) ~ "SSD",
      str_detect(memory, regex("flash", ignore_case = TRUE)) ~ "Flash Storage",
      str_detect(memory, regex("hdd", ignore_case = TRUE)) ~ "HDD",
      str_detect(memory, regex("hybrid", ignore_case = TRUE)) ~ "Hybrid",
      TRUE ~ NA_character_
    ),
    
    # Standardize operating system categories
    opsys = case_when(
      str_detect(opsys, regex("windows", ignore_case = TRUE)) ~ "Windows",
      str_detect(opsys, regex("mac", ignore_case = TRUE)) ~ "macOS",
      str_detect(opsys, regex("linux", ignore_case = TRUE)) ~ "Linux",
      str_detect(opsys, regex("chrome", ignore_case = TRUE)) ~ "Chrome OS",
      str_detect(opsys, regex("no os", ignore_case = TRUE)) ~ "No OS",
      TRUE ~ "Other"
    ),
    
    # Convert price to USD using an approximate exchange rate
    price_usd = round(price / 84)
  ) |>
  
  # Remove original price column after conversion
  select(-price)

# Reshape selected numeric specifications from wide to long format
tidy_data <- clean_data |>
  pivot_longer(
    cols = c(ram, weight, inches, storage_gb),
    names_to = "specification",
    values_to = "value"
  ) |>
  
  # Remove rows only when the pivoted value is missing
  drop_na(value)

# Inspect cleaned wide data
glimpse(clean_data)

# Inspect final tidy data
glimpse(tidy_data)

# View first few rows of tidy output
head(tidy_data)

```

##  Data Analysis: 

-   **Lowest average price** by brand

```{r}

# Top laptop brand affordability analysis

# Calculate average price for each laptop brand
brand_price <- clean_data |>
  filter(!is.na(company), !is.na(price_usd)) |>   # remove rows with missing brand or price
  group_by(company) |>
  summarise(
    avg_price = round(mean(price_usd), 2),        # average price per brand
    n_models = n(),                               # number of laptop models for each brand
    .groups = "drop"
  ) |>
  mutate(
    # Create affordability tiers based on average price
    affordability = case_when(
      avg_price < 400  ~ "Affordable (0–400 USD)",
      avg_price < 800  ~ "Mid-range (400–800 USD)",
      avg_price < 1200 ~ "Premium (800–1200 USD)",
      TRUE             ~ "High-end (1200+ USD)"
    ),

    # Order the tiers logically for plotting
    affordability = factor(
      affordability,
      levels = c(
        "Affordable (0–400 USD)",
        "Mid-range (400–800 USD)",
        "Premium (800–1200 USD)",
        "High-end (1200+ USD)"
      )
    )
  )

# Keep only brands with at least 5 models
brand_price_filtered <- brand_price |>
  filter(n_models >= 5)

# Inspect the result
glimpse(brand_price_filtered)
```

```{r}
ggplot(brand_price_filtered,
       aes(x = reorder(company, avg_price),
           y = avg_price,
           fill = affordability)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Laptop Brand Affordability by Average Price",
    subtitle = "Average laptop price by brand (USD)",
    x = "Laptop Brand",
    y = "Average Price (USD)",
    fill = "Affordability Tier"
  ) +
  theme_minimal()
```

## Conclusion

This analysis examined laptop brand affordability by calculating the average price of laptops for each manufacturer after converting prices to U.S. dollars. Brands were grouped into affordability tiers based on their average price, ranging from affordable to high-end.

The results show clear differences in market positioning across laptop brands. Mediacom and Acer fall into the affordable category with average prices below \$400. Brands such as HP, Lenovo, Asus, Dell, and Toshiba appear in the mid-range segment, with average prices between \$400 and \$800. Apple, Microsoft, MSI, and Samsung occupy the premium category, with average prices between \$800 and \$1200. Razer stands out as the only brand classified as high-end, with an average price exceeding \$2000.

These findings suggest that laptop manufacturers tend to occupy distinct price segments within the market. Some brands focus on budget-friendly devices, while others target premium or high-performance categories. Understanding these price tiers helps illustrate how different brands position their products within the broader laptop market.
