---
title: "Personalized Movie Recommender System"
author: "Madina Kudanova"
format: 
  html:
    code-fold: false
    toc: true
    toc-depth: 3
editor: visual
---

## Introduction

This assignment builds a **personalized recommendation system** using collaborative filtering. Unlike global baseline models that provide the same recommendations to everyone, this system learns individual user preferences to generate customized movie suggestions.

**Key Objectives:**

1.  Implement item-based collaborative filtering (IBCF)
2.  Generate personalized top-N recommendations
3.  Evaluate model performance using hold-out validation
4.  Compare predicted ratings against actual user ratings

## Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(recommenderlab)
library(dplyr)
library(ggplot2)
library(tidyr)

# Set seed for reproducibility
set.seed(123)


```

```{r load-data, message=FALSE}
# Load the movie ratings survey data
ratings_raw <- read.csv("https://raw.githubusercontent.com/MKudanova/Data607/refs/heads/main/w11/Movie%20Ratings%20Survey.csv")

# Display data structure
cat("Dataset dimensions:", dim(ratings_raw), "\n")
cat("Number of users:", nrow(ratings_raw), "\n")
cat("Number of movies:", ncol(ratings_raw) - 1, "\n\n")

head(ratings_raw[, 1:5])
```

## Data Preprocessing

```{r clean-data}
# Remove timestamp column
ratings_df <- ratings_raw[, -1]

# Extract movie names - they come after "...Masterpiece..."
colnames(ratings_df) <- gsub(".*Masterpiece\\.{3}(.*)", "\\1", colnames(ratings_df))

# Check result
cat("Cleaned column names:\n")
print(head(colnames(ratings_df)))
cat("\n")

# Convert to numeric matrix
ratings_matrix <- as.matrix(ratings_df)
mode(ratings_matrix) <- "numeric"

# Add user identifiers
rownames(ratings_matrix) <- paste0("User", 1:nrow(ratings_matrix))

# Check for data quality
cat("Missing ratings:", sum(is.na(ratings_matrix)), "\n")
cat("Total possible ratings:", nrow(ratings_matrix) * ncol(ratings_matrix), "\n")
cat("Sparsity:", round(sum(is.na(ratings_matrix)) / (nrow(ratings_matrix) * ncol(ratings_matrix)) * 100, 2), "%\n\n")

# Convert to recommenderlab format
ratings_rrm <- as(ratings_matrix, "realRatingMatrix")
ratings_rrm
```

### Data Exploration

```{r explore-data}
# Rating distribution
rating_values <- as.vector(ratings_matrix)
rating_values <- rating_values[!is.na(rating_values)]

cat("Rating statistics:\n")
cat("  Mean:", round(mean(rating_values), 2), "\n")
cat("  Median:", median(rating_values), "\n")
cat("  SD:", round(sd(rating_values), 2), "\n\n")

# Visualize rating distribution
ggplot(data.frame(rating = rating_values), aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Movie Ratings",
       x = "Rating", y = "Count") +
  theme_minimal()
```

## Model Training

### Train-Test Split

```{r split-data}
# Create evaluation scheme
# - 80% training, 20% testing
# - given = -1 means all but one item per test user is given
# - goodRating = 3 defines what counts as a "good" rating
scheme <- evaluationScheme(
  data = ratings_rrm,
  method = "split",
  train = 0.8,
  given = -1,
  goodRating = 3
)

# Extract datasets
train_set <- getData(scheme, "train")
known_set <- getData(scheme, "known")
unknown_set <- getData(scheme, "unknown")

cat("Training set:", dim(train_set), "\n")
cat("Test set (known):", dim(known_set), "\n")
cat("Test set (unknown):", dim(unknown_set), "\n")

```

### Build IBCF Model

```{r train-model}
# Train Item-Based Collaborative Filtering model
# k = 5: use 5 most similar items for prediction
# method = "cosine": use cosine similarity metric
ibcf_model <- Recommender(
  data = train_set,
  method = "IBCF",
  parameter = list(k = 5, method = "cosine")
)

cat("Model trained successfully!\n")
print(ibcf_model)
```

## Model Evaluation

### Generate Predictions

```{r predict}
# Predict ratings for test users
predicted_ratings <- predict(
  object = ibcf_model,
  newdata = known_set,
  type = "ratings"
)

# Convert to matrices for comparison
pred_matrix <- as(predicted_ratings, "matrix")
true_matrix <- as(unknown_set, "matrix")

# Verify alignment
cat("Prediction matrix dimensions:", dim(pred_matrix), "\n")
cat("True matrix dimensions:", dim(true_matrix), "\n")
cat("Dimensions match:", identical(dim(pred_matrix), dim(true_matrix)), "\n")
cat("Row names match:", identical(rownames(pred_matrix), rownames(true_matrix)), "\n")
cat("Column names match:", identical(colnames(pred_matrix), colnames(true_matrix)), "\n\n")
```

### Calculate Error Metrics

```{r evaluate}
# Extract matching predictions and actuals
# Only compare where we have both predicted and actual ratings
comparison_df <- data.frame(
  actual = as.vector(true_matrix),
  predicted = as.vector(pred_matrix)
) %>%
  filter(!is.na(actual), !is.na(predicted))

cat("Number of comparable predictions:", nrow(comparison_df), "\n\n")

# Calculate error metrics
rmse <- sqrt(mean((comparison_df$actual - comparison_df$predicted)^2))
mae <- mean(abs(comparison_df$actual - comparison_df$predicted))
mse <- mean((comparison_df$actual - comparison_df$predicted)^2)

# Display results
cat("=== MODEL PERFORMANCE ===\n")
cat("RMSE (Root Mean Squared Error):", round(rmse, 4), "\n")
cat("MAE  (Mean Absolute Error):    ", round(mae, 4), "\n")
cat("MSE  (Mean Squared Error):     ", round(mse, 4), "\n\n")

# Interpretation
cat("Interpretation:\n")
cat("- On average, predictions are off by", round(mae, 2), "rating points\n")
cat("- Lower values indicate better performance\n")
```

### Visualize Prediction Accuracy

```{r plot-predictions}
# Plot actual vs predicted ratings
ggplot(comparison_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  labs(title = "Actual vs Predicted Ratings",
       subtitle = paste("RMSE =", round(rmse, 3), "| MAE =", round(mae, 3)),
       x = "Actual Rating",
       y = "Predicted Rating") +
  theme_minimal() +
  coord_fixed(ratio = 1)

# Residual plot
comparison_df$residual <- comparison_df$actual - comparison_df$predicted

ggplot(comparison_df, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residual Plot",
       x = "Predicted Rating",
       y = "Residual (Actual - Predicted)") +
  theme_minimal()
```

## Generate Recommendations

### Top-N Recommendations

```{r top-n-recommendations}
# Generate top-3 movie recommendations for each user
top_n <- 3

top_recommendations <- predict(
  object = ibcf_model,
  newdata = ratings_rrm,
  type = "topNList",
  n = top_n
)

# Convert to readable format
recommendation_list <- as(top_recommendations, "list")

# Display recommendations for first 10 users
cat("TOP", " RECOMMENDATIONS\n\n", sep = "")
for (i in 1:min(10, length(recommendation_list))) {
  user_name <- names(recommendation_list)[i]
  movies <- recommendation_list[[i]]
  
  cat("User:", user_name, "\n")
  if (length(movies) > 0) {
    cat("  Recommended:", paste(movies, collapse = ", "), "\n\n")
  } else {
    cat("  No recommendations available\n\n")
  }
}

# Summary statistics
rec_lengths <- sapply(recommendation_list, length)
cat("Recommendation summary:\n")
cat("  Users with", top_n, "recommendations:", sum(rec_lengths == top_n), "\n")
cat("  Users with < ", top_n, "recommendations:", sum(rec_lengths < top_n), "\n")
cat("  Users with 0 recommendations:", sum(rec_lengths == 0), "\n")
```

### Create Recommendation Table

```{r recommendation-table}
# Create a tidy table of all recommendations
recommendations_tidy <- data.frame(
  user = rep(names(recommendation_list), sapply(recommendation_list, length)),
  movie = unlist(recommendation_list),
  rank = unlist(lapply(recommendation_list, seq_along))
) %>%
  pivot_wider(names_from = rank, values_from = movie, names_prefix = "Rec_")

head(recommendations_tidy, 10)
```

## Model Insights

### Item Similarity Matrix

```{r similarity-matrix}
# Extract the learned item-item similarity matrix
similarity_matrix <- getModel(ibcf_model)$sim

# Convert to regular matrix for inspection
sim_matrix_view <- as.matrix(similarity_matrix)

# Display sample of similarities
cat("Sample of item-item similarities:\n\n")
print(round(sim_matrix_view[1:min(6, nrow(sim_matrix_view)),
                             1:min(6, ncol(sim_matrix_view))], 3))

# Find most similar movie pairs
if (ncol(sim_matrix_view) >= 2) {
  # Get upper triangle (avoid duplicates)
  sim_matrix_upper <- sim_matrix_view
  sim_matrix_upper[lower.tri(sim_matrix_upper, diag = TRUE)] <- NA
  
  # Find top 5 most similar pairs
  sim_df <- as.data.frame(as.table(sim_matrix_upper)) %>%
    filter(!is.na(Freq)) %>%
    arrange(desc(Freq)) %>%
    head(5)
  
  cat("\n\nMost similar movie pairs:\n")
  for (i in 1:nrow(sim_df)) {
    cat(i, ". ", as.character(sim_df$Var1[i]), " <-> ", 
        as.character(sim_df$Var2[i]), " (similarity: ", 
        round(sim_df$Freq[i], 3), ")\n", sep = "")
  }
}
```

## Summary

### Key Findings

```{r summary}
cat("MODEL SUMMARY\n\n")
cat("Algorithm: Item-Based Collaborative Filtering (IBCF)\n")
cat("Training approach: 80/20 hold-out validation\n")
cat("Similarity metric: Cosine similarity\n")
cat("Neighborhood size (k): 5\n\n")

cat("Performance Metrics:\n")
cat("  RMSE:", round(rmse, 4), "\n")
cat("  MAE: ", round(mae, 4), "\n\n")

cat("Output:\n")
cat("  Top-N recommendations per user: ", top_n, "\n")
cat("  Total users:", nrow(ratings_matrix), "\n")
cat("  Total movies:", ncol(ratings_matrix), "\n")
```

### **Conclusion**

This project successfully implemented a personalized movie recommendation system using Item-Based Collaborative Filtering (IBCF), which generates customized suggestions for each user based on their individual rating patterns—a significant improvement over non-personalized baseline models.

I chose IBCF because it calculates similarities between movies based on user ratings, then recommends new movies similar to ones a user already liked. With only 26 users but consistent movie ratings, an item-based approach is more stable than user-based filtering. The model uses cosine similarity and considers the 5 most similar items when making predictions

Using 80/20 hold-out validation, the model achieved an RMSE of 2.011 and MAE of 1.60, meaning predictions are off by approximately 1.6 rating points on average. The actual vs. predicted ratings plot shows positive correlation, indicating the model successfully learned user preferences.

The system generates top-3 personalized movie recommendations per user, though coverage varied based on how many movies each user had already rated. The diversity in recommendations—different users receiving different suggestions—confirms genuine personalization. The similarity matrix revealed interesting patterns, such as Forrest Gump and The Sound of Music being most alike based on user ratings, which directly drives the recommendation logic.

Main limitations include the cold start problem, small dataset size (6 movies, 26 users), and rating sparsity. Future improvements could include testing different parameters, implementing cross-validation, trying matrix factorization algorithms, and adding diversity metrics.

Overall, this IBCF model successfully delivers personalized recommendations by learning from collective user behavior, representing a meaningful advancement over non-personalized approaches.
